The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-05-29 source/sec_strongly_convex_smooth.tex
update figures
updated strongly convex smooth section

In this section, we show that FedAvg with SGD and Nesterov Accelerated SGD has $O(1/NT)$ rate under $\mu$-strong
convexity and $L$-smoothness. The result improves on previous results of
$O(1/T)$ with a linear speedup in the number of devices $N$.

\subsection{SGD}

\begin{theorem}
	Suppose $F_{k}$ is $L$-smooth and $\mu$-strongly convex for all
	$k$. Let $\kappa=\frac{L}{\mu}$, $\gamma=\max\{8\kappa-1,E\}$ where
	$E$ is the communication interval, and learning rates 
	\begin{align*}
	\alpha_{t} & =\frac{c}{\mu(\gamma+t)}
	\end{align*}
	so that $\alpha_{t}\leq2\alpha_{t+E}$, and where $0\leq c\leq1$
	is small enough such that
	$\alpha_{t} \leq\frac{1}{4L}$ and 
	$\alpha_{t} \leq\frac{1}{N}$ for all $t\geq0$.

	Then with full device participation, if $\nu_{max}=\max_{k}\frac{1}{N}p_{k}$,
	we have 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq C\frac{\kappa}{\gamma+T}\cdot(8E^{2}LG^{2}/T+\frac{1}{N}\nu_{max}^{2}\sigma^{2})
	\end{align*}
	so that as long as $E=O(\sqrt{\frac{N}{T}})$, 
	\begin{align*}
	\mathbb{E}F(w_{T})-F^{\ast} & \leq C(8LG^{2}+\nu_{max}^{2}\sigma^{2})\frac{\kappa}{N(\gamma+T)}
	\end{align*}
	and with partial participation with $K$ sampled devices at each
	communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq C\frac{\kappa}{\gamma+T}\cdot(8E^{2}LG^{2}/T+\frac{1}{N}\nu_{max}^{2}\sigma^{2}+\frac{4}{K}E^{2}G^{2})
	\end{align*}
	so that if $E=O(1)$, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast} & \leq C(4E^{2}G^{2}+\nu_{max}^{2}\sigma^{2})\frac{\kappa}{K(\gamma+T)}+\frac{8E^{2}LG^{2}}{T^{2}}
	\end{align*}
\end{theorem}
\begin{proof}
	See the appendix.
\end{proof}

The convergence result implies that with fixed $E$, as long as $N$ satisfies $E=O(\sqrt{T/N})$, there is linear speedup. When $N$ exceeds an upper bound, however, linear speedup may fail to happen, and $T/N$ may remain constant, i.e. the number of iterations required for convergence may increase with $N$.

>>>>

In this section, we show that FedAvg with local SGD has $O(1/NT)$ convergence
rate for $\mu$-strongly convex and $L$-smooth objectives. The result
improves on the $O(1/T)$ result of {[}ICLR{]} with a linear speedup
in the number of devices $N$. Moreover, it makes a distinction on
the choice of $E$ with partial and full participation. With full
participation, $E$ can be chosen as large as $O(\sqrt{\frac{T}{N}})$
without degrading the linear speedup in the number of workers. On
the other hand, with partial participation, $E$ must be $O(1)$. 
\begin{theorem}
	Suppose $F_{k}$ is $L$-smooth and $\mu$-strongly convex for all
	$k$. Let $\kappa=\frac{L}{\mu}$, $\gamma=\max\{8\kappa-1,E\}$ where
	$E$ is the communication delay, and learning rates 
	\begin{align*}
	\alpha_{t} & =\frac{c}{\mu(\gamma+t)}
	\end{align*}
	so that $\alpha_{t}\leq2\alpha_{t+E}$, and where $0\leq c\leq1$
	is such that $\alpha_{t}\leq\frac{1}{8L}$ for all $t\geq0$.

	Then with full device participation, if $\nu_{max}=\max_{k}\frac{1}{N}p_{k}$,
	we have 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq C\cdot(\frac{\kappa\nu_{max}^{2}\sigma^{2}}{NT}+\frac{5\kappa E^{2}LG^{2}}{T^{2}})
	\end{align*}
	and with partial participation with $K$ sampled devices at each
	communication round, 
	\begin{align*}
	\mathbb{E}F(\overline{w}_{T})-F^{\ast}\leq C(\frac{\kappa\nu_{max}^{2}\sigma^{2}}{NT}+\frac{4\kappa E^{2}G^{2}}{KT}+\frac{5\kappa E^{2}LG^{2}}{T^{2}})
	\end{align*}
\end{theorem}
%
\begin{remark}
	We remark on the linear speedup and the choice of $E$. In full participation,
	the leading term is $\frac{1}{NT}\nu_{max}^{2}\sigma^{2}$ whereas
	$\frac{5\kappa E^{2}LG^{2}}{T^{2}}$ is a lower order term if $E$
	is chosen constant. This implies that $E$ can be chosen to be as
	large as $E=O(\sqrt{\frac{T}{N}})$ which results in $\frac{5\kappa E^{2}LG^{2}}{T^{2}}=O(\frac{1}{NT})$,
	which will be on the same order as the leading term. This is in contrast
	with the convergence rate in {[}ICLR{]} where $E$ cannot scale with
	$T$ and must be $O(1)$, even with full participation. Thus our bound
	yields a more efficient communication complexity than previous analyses.
	We note however that this is only possible when there is full participation,
	and when $\nu_{max}$ is close to 1. When there is partial participation,
	however, $E$ has to stay $O(1)$, as the term leading term in the
	bound for partial participation contains $\frac{4\kappa E^{2}G^{2}}{KT}$.
	Thus our result hightlights the difference between the communication
	efficiency of full and partial participation when an $O(1/NT)$ bound
	on the optimality gap is desired.

	\textbf{}%
	\begin{comment}
	This implies that $E$ canot be chosen $O(T^{\beta})$ for any $\beta>0$
	without degrading the performance. This should be checked in experiments,
	whether with partial participation if the communication round is set
	to scale with $T$, the convergence deteriorates. This is in constrast
	with the full participation case, where $E=O(\sqrt{\frac{T}{N}})$
	is allowed. \textbf{If we can confirm that full participation allows
	linear speedup with $\nu=N\cdot\max_{k}p_{k}\approx1$ and $E=O(\sqrt{\frac{T}{N}})$,
	whereas partial participation only allows $E=O(1)$, then this would
	an interesting phenomenon that is not reported by previous studies!}
	\end{comment}
\end{remark}
